---
author: "Adrian Wong & Zijian Wu"
title: "The Effects of Chemical Properties of Red and White Wine on Quality"
date: "2026-03-05"
output:
  pdf_document:
    number_sections: true
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r echo=FALSE, include=FALSE, message=FALSE}
# all library load statements here
library(caret) # for training and cross validation
library(knitr) # for kable table rendering
library (tidyverse) # tibble
library(ggplot2) # plotting
library(car) # for VIF
# two libraries for applying robust std errors:
library(lmtest) # for Breusch-Pagan test
library(sandwich)
library(modelsummary)

r.version <- "4.3.2"
seed.val <- 123456
```

Notes: This is an Rmd file template for the project. You may edit the
headings below as fits your analysis. You may add headings and
sub-headings as well. Please read the requirements for project 1
milestone 1 and for the final submission.

For the Overview: A brief intro to the domain to give context and your
main question go in this space. You don't need a heading for the
overview (unless you want one).

There are many different factors on what determines the quality of a
wine.

The main research question of this analysis is how do the chemical
properties of wine affect the expert rating of quality of wine?

# Data Description and Processing

Variables:\
fixed_acidity: volatile_acidity: citric_acid: residual_sugar: chlorides:
free_sulfurdioxide: total_sulfurdioxide: density: pH: sulphates:
alcohol:

Outcome:\
quality:

The data set was checked for missing values. Of the original 6,497
observations, we removed 5 observations that contained missing values
leaving 6,492 complete observations.

We used the quality as an outcome.

```{r echo=FALSE, include=FALSE}
wine.df <- read.csv("dataset4.csv")
wine.df <- na.omit(wine.df)
wine.df$quality <- as.numeric(wine.df$quality)
```

# Data Exploration

```{r echo=FALSE, out.width="70%", out.height="35%", fig.align = 'center', fig.cap="Wine Quality index distribution."}
hist(wine.df$quality,
main = "", # turn off the title as using the fig.cap
xlab = "Index Value",
ylab = "Frequency")
```

```{r echo=FALSE}
# table of descriptive stats for the performance index scores
statistic <- c("Min", "2nd Quant. (25%)", "Mean","Median", "4th Quant. (75%)", "Max", "StdDev")
data<-wine.df$quality
quantiles<-quantile(data)
# extracting the number with as.numeric because quantiles[2] is a named number.
second.quant<-as.numeric(quantiles[2])
fourth.quant<-as.numeric(quantiles[4])
mn<-min(data)
mx<-max(data)
avg<-round(mean(data), 2)
md<-median(data)
std<-round(sd(data), 2)
value <- c(mn,second.quant,avg,md,fourth.quant,mx,std)
st.tbl <- cbind(statistic, value)
kable(st.tbl, caption="Wine Quality descriptive statistics")
```

# Modeling

## Hypothesis

## Model 1:
```{r}
fit1 <- lm (quality ~ ., data=wine.df)
# summary object useful for values from fit.
fit1.summary<-summary(fit1)
coef.vals <- coef(fit1.summary)
kable(coef.vals, caption="Summary of coefficients for model 1.")


fit1.summary
```



```{r}
vif <-vif(fit1)
vif

```


## Model 2:

```{r}
fit2 <- lm(quality~volatile_acidity+citric_acid+chlorides+free_sulfurdioxide+total_sulfurdioxide+pH+sulphates,data=wine.df)
fit2.summary<-summary(fit2)
coef.vals <- coef(fit2.summary)
kable(coef.vals, caption="Summary of coefficients for model 2.")
```

```{r}
bptest <-bptest(fit2)
bptest
```


```{r echo=FALSE, out.width="80%", out.height="35%", fig.align = 'center', fig.cap="Plot of of residuals vs. fitted values and Q-Q plot for model 1."}

par(mfrow = c(1,2))
# residuals vs fitted
plot(fit2, which = 1)
#QQ plot
plot(fit2, which = 2)
par(mfrow = c(1,1)) # reset to default window
```


# Predicting with Models 1 and 2


```{r echo=FALSE}
train.control <- trainControl(method = "repeatedcv", number = 10, repeats= 3, p = 0.80) 

RNGversion(r.version)
set.seed(seed.val)
fit1.train <- train(performance_index ~ ., data = data.df, method = "lm", trControl = train.control)

fit2.train <- train(performance_index ~ study_hrs+prior_score+practice, data = data.df, method = "lm", trControl = train.control)
```

```{r echo=FALSE}
label <- c("Model 1", "Model 2")
cols<-c("Model", "Avg RMSE","Std Dev")
# get the RMSE vectors
rmse.one <- fit1.train$resample$RMSE
rmse.two <- fit2.train$resample$RMSE

avg.rmse1<-round(mean(rmse.one),2)
avg.rmse2<-round(mean(rmse.two),2)

sd.rmse1<-round(sd(rmse.one),4)
sd.rmse2<-round(sd(rmse.two),4)

rmse.vals <- c(avg.rmse1, avg.rmse2)
sd.vals <- c(sd.rmse1, sd.rmse2)

imb.tbl <- cbind(label, rmse.vals, sd.vals)
kable(imb.tbl, col.names=cols, caption="Mean RMSE and standard deviation for models 1 and 2.")
```


```{r echo=FALSE, out.width="75%", out.height="40%", fig.align = 'center', fig.cap="RMSE over 3x10=30 folds for models 1 and 2."}
num.folds <- 30
plot(rmse.one, 
     xlab = "fold",
     xaxp = c(1, num.folds, num.folds-1),
     ylab = "rmse",
     ylim=c(min(rmse.one,rmse.two), max(rmse.one,rmse.two)),
     col = "blue")
lines(rmse.one, col = "blue")
points(rmse.two, col="red")
lines(rmse.two, col = "red",lty = 2)

legend("bottomleft", legend=c("model 1", "model 2"),
       col=c("blue", "red"), lty=1:2, cex=0.5)
```




# Summary and Conclusions
