---
author: "Adrian Wong & Zijian Wu"
title: "The Effects of Chemical Properties of Red and White Wine on Quality"
date: "2026-03-05"
output:
  pdf_document:
    number_sections: true
editor_options: 
  chunk_output_type: console
---
```{r echo=FALSE, include=FALSE, message=FALSE}
# all library load statements here
library(caret) # for training and cross validation
library(knitr) # for kable table rendering
library (tidyverse) # tibble
library(ggplot2) # plotting
library(car) # for VIF
# two libraries for applying robust std errors:
library(lmtest) # for Breusch-Pagan test
library(sandwich)
library(modelsummary)

r.version <- "4.3.2"
seed.val <- 123456
```

Notes: This is an Rmd file template for the project. You may edit the headings below as fits your analysis. You may add headings and sub-headings as well. Please read the requirements for project 1 milestone 1 and for the final submission.

For the Overview: A brief intro to the domain to give context and your main question go in this space. You don't need a heading for the overview (unless you want one).

There are many different factors on what determines the quality of a wine. 

The main research question of this analysis is how do the chemical properties of wine affect the expert rating of quality of wine?

# Data Description and Processing

Variables: \
fixed_acidity: 
volatile_acidity: 
citric_acid:
residual_sugar:
chlorides:
free_sulfurdioxide:
total_sulfurdioxide:
density:
pH:
sulphates:
alcohol:

Outcome:\
quality:


The data set was checked for missing values. Of the original 6,497 observations, we removed 5 observations that contained missing values leaving 6,492 complete observations.

We used the quality as an outcome.

```{r echo=FALSE, include=FALSE}
wine.df <- read.csv("dataset4.csv")
wine.df <- na.omit(wine.df)
wine.df$quality <- as.numeric(wine.df$quality)
```

# Data Exploration

```{r echo=FALSE, out.width="70%", out.height="35%", fig.align = 'center', fig.cap="Wine Quality index distribution."}
hist(wine.df$quality,
main = "", # turn off the title as using the fig.cap
xlab = "Index Value",
ylab = "Frequency")
```

```{r echo=FALSE}
# table of descriptive stats for the performance index scores
statistic <- c("Min", "2nd Quant. (25%)", "Mean","Median", "4th Quant. (75%)", "Max", "StdDev")
data<-wine.df$quality
quantiles<-quantile(data)
# extracting the number with as.numeric because quantiles[2] is a named number.
second.quant<-as.numeric(quantiles[2])
fourth.quant<-as.numeric(quantiles[4])
mn<-min(data)
mx<-max(data)
avg<-round(mean(data), 2)
md<-median(data)
std<-round(sd(data), 2)
value <- c(mn,second.quant,avg,md,fourth.quant,mx,std)
st.tbl <- cbind(statistic, value)
kable(st.tbl, caption="Wine Quality descriptive statistics")
```

# Modeling 


## Hypothesis


## Model 1:


## Model 2: 


# Predicting with Models 1 and 2


# Summary and Conclusions

